Constructing Biological Regulatory Networks from Process Hitting models

### 2013-07-23 parts/example.tex
TODOs
cleanup

\todo{Proofread this section}

\updatedmax{%
This section details the implementation and usage of the piece of software developed for this work.
\pref{ssec:appli} gives a practical and detailed application of our results on a biological model taken from the literature,
while \pref{ssec:cpu} gives data related to the computation times of our software when applied to several large biological models.
}

@@

\updatedmax{%
\subsection{Biological application}\label{ssec:appli}

@@

\begin{table}
\begin{center}
  \begin{tabular}{r|c|c|c|c}
    \textbf{Model} & \textbf{Components} & \textbf{\# CS} & \textbf{$\Delta t$ IG} & \textbf{$\Delta t$ K}
\\\hline\hline
    ERBB-G1/S \cite{Sahin09} & $20$ & $22$ & <1s & <1s
\\\hline
    T cell receptor \cite{Klamt06} & $40$ & $14$ & <1s & <1s
\\\hline
    T cell receptor \cite{SaezRodriguez2007} & $94$ & $39$ & 10s & ---
\\\hline
    EGFR/ErbB signaling \cite{Samaga2009} & $104$ & $89$ & 3min 20s & ---
  \end{tabular}
\end{center}
  \caption{Other models \rewrite{Caption}
  Computation times and several pieces of data related to the IG and parametrization inferences of fours biological models.
  The third column gives the number of cooperative sorts used in the models to model joint actions,
  and the fourth (resp.~fifth) column gives the computation times of the IG inference (resp.~the parametrization inference).
  }
  \label{tb:computation}
\end{table}
}%

\begin{comment}
\subsection{\warnul{Old version of this section}}

The inference method described in this paper has been implemented as part of
\textsc{Pint}\footnote{Available at \url{http://process.hitting.free.fr}}, which gathers PH related
tools.
Our implementation mainly consists in ASP programs that are solved using Clingo\footnote{Available
at \url{http://potassco.sourceforge.net}}.
The IG and parameters inference can be performed using the command
\texttt{ph2thomas -i model.ph -{}-dot ig.dot}
where \texttt{model.ph} is the PH model in \textsc{Pint} format and \texttt{ig.dot} is an output of the inferred IG in DOT format.
The (possibly partial) inferred parametrization will be returned on the standard output.
The admissible parametrizations enumeration is performed when adding the \texttt{-{}-enumerate}
parameter to the command.

Applied to the example in \pref{fig:runningPH-2} where cooperations have been defined,
our method infers the IG and parametrization given in \pref{fig:runningBRN}.
Regarding the example in \pref{fig:runningPH-1}, the same IG is inferred, as well as for the
parametrization except for the parameters $K_{a,\{b\},\{c\}}$ and $K_{a,\{c\},\{b\}}$ which are
undefined (because of the lack of cooperativity between $b$ and $c$).
In such a case, this partial parametrization allows 36 admissible complete parametrizations, as two
parameters with 3 potential values could not be inferred.
If we constrain these latter parameters so that they contain exactly one element, we obtain only 9
admissible parametrizations.

The current implementation can successfully handle large PH models of BRNs found in the literature
such as an ERBB receptor-regulated G1/S transition model from \cite{Sahin09} which contains 20
components and 15 cooperative sorts, and a T-cells receptor model from \cite{Klamt06} which contains 40
components and 14 cooperative sorts\footnote{Both models are available as examples distributed with \textsc{Pint}.}.
For each model, IG and parameters inferences are performed together in less than a second
on a standard desktop computer.
%\footnote{Using a Dell Inspiron 1720 laptop, with an Intel Core 2 Duo CPU T5550 ($2 \times 1.83\text{GHz}$)
%and 3.9 Gib memory, on an Ubuntu 11.10 64-bits OS}
After removing the cooperations from these models (leaving only raw actions), the inferences allow to
determine 40 parameters out of 195 for the 20 components model, and 77 out of 143 for the 40 components model.
As we thus have an order of magnitude of respectively $10^{31}$ and $10^{73}$ admissible parametrizations,
these models would therefore be more efficiently studied as PH than as BRNs.
We note that the complexity of the method is exponential in the number of regulators of one
component and linear in the number of components.
% We note that despite its smaller size in term of components, the ERBB transmission model takes more time to be computed because the biggest cooperative sorts contain more processes (up to 32 processes) than in the T-cells receptor model (up to 8 processes).

A PH model can be built based on information found in the literature about the local influences between components.
The precision of this knowledge will determine the precision of the modeled activations and inhibitions,
and some information is likely to help in the representation of cooperations.
\end{comment}

>>>>

%\todo{Proofread this section}

This section details the implementation and usage of the piece of software developed for this work.
\pref{ssec:appli} gives a practical and detailed application of our results on a biological model taken from the literature,
while \pref{ssec:cpu} gives data related to the computation times of our software when applied to several large biological models.

@@

\subsection{Biological application}\label{ssec:appli}

@@

\begin{table}
\begin{center}
  \begin{tabular}{r|c|c|c|c}
    \textbf{Model} & \textbf{Components} & \textbf{\# CS} & \textbf{$\Delta t$ IG} & \textbf{$\Delta t$ K}
\\\hline\hline
    ERBB-G1/S \cite{Sahin09} & $20$ & $22$ & <1s & <1s
\\\hline
    T cell receptor \cite{Klamt06} & $40$ & $14$ & <1s & <1s
\\\hline
    T cell receptor \cite{SaezRodriguez2007} & $94$ & $39$ & 10s & ---
\\\hline
    EGFR/ErbB signaling \cite{Samaga2009} & $104$ & $89$ & 3min 20s & ---
  \end{tabular}
\end{center}
\caption{
  Computation times and several pieces of data related to the IG and parametrization inferences of fours biological models.
  The third column gives the number of cooperative sorts used in the models to model joint actions,
  and the fourth (resp.~fifth) column gives the computation times of the IG inference (resp.~the parametrization inference).
  }
\label{tb:computation}
\end{table}

### 2013-07-23 parts/infer-ig.tex
cleanup

\updated{%
The Interaction Graph (IG) is an abstract representation of the direct qualitative influences,
positive and/or negative, between the components of the system.
As discussed in \pref{sec:intro}, the IG allows to efficiently characterise global dynamical
properties for the concrete system, such as the capability for multi-stationarity or oscillation.}

\updated{%
In a typical biological network modelling process, a prior IG is generally the starting point for
the formal system specification.
However, it is common that the prior IG actually refers to interactions that reveal to be 
non-effective with respect to the dynamics.
Hence, deriving the IG directly from the dynamical models lead to more concise IGs, enhancing the
conclusiveness of static analyses based upon this abstract representation.}

\updated{%
In this section, we formally derive the IG corresponding to a given PH that is well-formed for BRN
modelling.}
%
%In order to infer a complete BRN, one has to find the Interaction Graph (IG) first, as some
%constraints on the parametrization rely on it.
%Inferring the IG is an abstraction step which consists in determining the global influence of
%components on each of its successors.
%
%
This section first introduces the notion of focal processes within a PH (\pref{ssec:focal})
which is used to characterize well-formed PH for IG inference (\pref{ssec:wf}).
%and as well used by the parametrization inference presented in \pref{sec:infer-K}.
The rules for inferring the interactions between components from a PH are
described in \pref{ssec:infer-IG}.
We consider hereafter a global PH $(\PHs,\PHl,\PHa)$ on which the IG inference is to be
performed.

>>>>

The Interaction Graph (IG) is an abstract representation of the direct qualitative influences,
positive and/or negative, between the components of the system.
As discussed in \pref{sec:intro}, the IG allows to efficiently characterise global dynamical
properties for the concrete system, such as the capability for multi-stationarity or oscillation.

In a typical biological network modelling process, a prior IG is generally the starting point for
the formal system specification.
However, it is common that the prior IG actually refers to interactions that reveal to be 
non-effective with respect to the dynamics.
Hence, deriving the IG directly from the dynamical models lead to more concise IGs, enhancing the
conclusiveness of static analyses based upon this abstract representation.

In this section, we formally derive the IG corresponding to a given PH that is well-formed for BRN
modelling.
%
%In order to infer a complete BRN, one has to find the Interaction Graph (IG) first, as some
%constraints on the parametrization rely on it.
%Inferring the IG is an abstraction step which consists in determining the global influence of
%components on each of its successors.
%
%
This section first introduces the notion of focal processes within a PH (\pref{ssec:focal})
which is used to characterize well-formed PH for IG inference (\pref{ssec:wf}).
%and as well used by the parametrization inference presented in \pref{sec:infer-K}.
The rules for inferring the interactions between components from a PH are
described in \pref{ssec:infer-IG}.
We consider hereafter a global PH $(\PHs,\PHl,\PHa)$ on which the IG inference is to be
performed.

### 2013-07-23 parts/intro.tex
cleanup

While the formal checking of dynamical properties is often limited to small networks because of the
state graph explosion, \updated{an other major} drawback of this framework is the difficulty to
specify Thomas' parameters, especially for large networks.

In order to address the formal analysis of dynamical properties within very large BRNs, we recently
introduced in \cite{PMR10-TCSB} a new formalism, named the \emph{Process Hitting} (PH), to model
concurrent systems having components with a few qualitative levels.
A PH model describes, in an atomic manner, the possible evolutions of a process (representing one
component at one level) triggered by the hit of at most one other process in the system.
This framework can be seen as a special class of formalisms like Petri Nets or Communicating Finite
State Machines, \updated{where the arity and the effect of synchronisations between components is
restricted.}
Thanks to the particular structure of interactions within a PH, very efficient static analysis
methods have been developed to over- and under-approximate reachability properties making tractable
the formal analysis of the qualitative dynamics of BRNs with hundreds or thousands of components,
\updated{which was impossible with other state-of-the-art approaches \cite{PMR12-MSCS,PAK13-CAV,FPMR13-CS2Bio}}.

@@

\updated{%
The aim of this paper is to formally establish links between several layers suited for specifying the
qualitative dynamics of BRNs, namely:
the Interaction Graph (IG), referencing the sign of direct regulations between components;
the Process Hitting (PH), which permits to specify the qualitative dynamics of the system with
various degrees of knowledge for joint regulations;
and Thomas' parameters, which completely specify the functions governing the evolution of
components.}
\updatedmax{%
Another motivation for the work depicted in this paper is that it constitutes an important step
to go further in the study of large biological regulatory networks systems beyond simple analysis:
it makes it possible to partly control some behaviors.
Indeed, when one wants to modify some behaviors
(e.g.~so as to avoid the reachability of some states or to create stable states)
we intend to be able to get upstream the results in accordance at the regulatory level.
The outcome of the present work is that we could make such changes on the PH model
and then recover afterwards the BRN corresponding to the transformed behavioral system.
In a more general framework and in an interesting way,
we should be able to synthesize families of BRN having some featuring behavioral properties.
}

\updated{%
Firstly, we derive rules that, given a PH model, infer the IG corresponding to the dynamics of the
encoded BRN.
The obtained IG relates only components that have actual influence in the dynamics.
This typically results into IGs that contain less interaction than an hypothetical starting prior
IG, as the specification of component evolutions may reveal non-functional regulations.
Based on the derived IG, various static analysis allows to conclude on global properties
of the system dynamics (see \cite{PR11-SASB} for a short survey).
The most known are the Thomas' conjectures (that have been proved since,
e.g.~\cite{RRT08,RiCo07,Richard2010378}
for Boolean and discrete networks),
which relate the absence of
positive cycle to the impossibility for multi-stationarity (distinct attractors),
and the absence of negative cycle to the impossibility for oscillatory behaviours.
}

\updated{%
Second, we tackle the characterisation of Thomas' parameters that are compatible with a given PH
model, \ie that fully specify the evolution of components while respecting the cooperations
specified in PH.
More formally, the resulting BRN dynamics is ensured to be included in the PH dynamics: any
transition in the (asynchronous) Thomas model exists in the PH model.
In practice, the number of compatible Thomas parameters can be huge for large networks where only a
few cooperations are specified.
This highlights the fact that the PH is very suited to analyse dynamics of large-scale networks
where the knowledge on cooperations is often incomplete:
instead of dealing with a possibly non-tractable number of Thomas models, one can already capture
precise dynamical features using a single PH models which gathers all the plausible dynamics.
}

\updated{%
Finally, we discuss on an implementation of these inference schemes using Answer Set
Programming (ASP) \cite{Baral03},  which turns out to be effective for these enumerative searches.
}%
\updatedmax{%
The whole method is also applied to the biological model of ERBB receptor-regulated G1/S transition,
containing 20 components,
which allows to detail how joint actions can be defined inside a PH model in order
to refine its dynamics, and study a more precise class of underlying Thomas models.
}
%\todo{describe application}

%The outcome of this work is twofold.
%The first benefit is that such an approach makes it possible to refine the construction of
%BRNs with a partial and progressively brought knowledge in PH, while being able to export such
%models in the Thomas' framework.
%This work thus strengthens the formal link between both modelings.
%The second feature of our method is that it can be applied on very large BRNs.

@@

\updated{%
The contributions presented in this paper significantly extend and improve the preliminary results
presented in \cite{FPIMR12-CMSB}.
In addition of improving the efficiency and accuracy of the IG inference, we have added support for
unsigned regulations, that are regulations being positive or negative depending on a particular
context.
This allows to apply our methodology to a wider class of models.
}%
\updatedmax{%
Furthermore, some parts have been completely rewritten in order to explain more precisely the efficiency of the ASP implementation,
and give a new detailed biological application of our work.
}%

>>>>

While the formal checking of dynamical properties is often limited to small networks because of the
state graph explosion, an other major drawback of this framework is the difficulty to
specify Thomas' parameters, especially for large networks.

In order to address the formal analysis of dynamical properties within very large BRNs, we recently
introduced in \cite{PMR10-TCSB} a new formalism, named the \emph{Process Hitting} (PH), to model
concurrent systems having components with a few qualitative levels.
A PH model describes, in an atomic manner, the possible evolutions of a process (representing one
component at one level) triggered by the hit of at most one other process in the system.
This framework can be seen as a special class of formalisms like Petri Nets or Communicating Finite
State Machines, where the arity and the effect of synchronizations between components is
restricted.
Thanks to the particular structure of interactions within a PH, very efficient static analysis
methods have been developed to over- and under-approximate reachability properties making tractable
the formal analysis of the qualitative dynamics of BRNs with hundreds or thousands of components,
which was impossible with other state-of-the-art approaches \cite{PMR12-MSCS,PAK13-CAV,FPMR13-CS2Bio}.

@@

The aim of this paper is to formally establish links between several layers suited for specifying the
qualitative dynamics of BRNs, namely:
the Interaction Graph (IG), referencing the sign of direct regulations between components;
the Process Hitting (PH), which permits to specify the qualitative dynamics of the system with
various degrees of knowledge for joint regulations;
and Thomas' parameters, which completely specify the functions governing the evolution of
components.
Another motivation for the work depicted in this paper is that it constitutes an important step
to go further in the study of large biological regulatory networks systems beyond simple analysis:
it makes it possible to partly control some behaviors.
Indeed, when one wants to modify some behaviors
(e.g.~so as to avoid the reachability of some states or to create stable states)
we intend to be able to get upstream the results in accordance at the regulatory level.
The outcome of the present work is that we could make such changes on the PH model
and then recover afterwards the BRN corresponding to the transformed behavioral system.
In a more general framework and in an interesting way,
we should be able to synthesize families of BRN having some featuring behavioral properties.

Firstly, we derive rules that, given a PH model, infer the IG corresponding to the dynamics of the
encoded BRN.
The obtained IG relates only components that have actual influence in the dynamics.
This typically results into IGs that contain less interaction than an hypothetical starting prior
IG, as the specification of component evolutions may reveal non-functional regulations.
Based on the derived IG, various static analysis allows to conclude on global properties
of the system dynamics (see \cite{PR11-SASB} for a short survey).
The most known are the Thomas' conjectures (that have been proved since,
e.g.~\cite{RRT08,RiCo07,Richard2010378}
for Boolean and discrete networks),
which relate the absence of
positive cycle to the impossibility for multi-stationarity (distinct attractors),
and the absence of negative cycle to the impossibility for oscillatory behaviours.

Second, we tackle the characterisation of Thomas' parameters that are compatible with a given PH
model, \ie that fully specify the evolution of components while respecting the cooperations
specified in PH.
More formally, the resulting BRN dynamics is ensured to be included in the PH dynamics: any
transition in the (asynchronous) Thomas model exists in the PH model.
In practice, the number of compatible Thomas parameters can be huge for large networks where only a
few cooperations are specified.
This highlights the fact that the PH is very suited to analyse dynamics of large-scale networks
where the knowledge on cooperations is often incomplete:
instead of dealing with a possibly non-tractable number of Thomas models, one can already capture
precise dynamical features using a single PH models which gathers all the plausible dynamics.

Finally, we discuss on an implementation of these inference schemes using Answer Set
Programming (ASP) \cite{Baral03},  which turns out to be effective for these enumerative searches.
The whole method is also applied to the biological model of ERBB receptor-regulated G1/S transition,
containing 20 components,
which allows to detail how joint actions can be defined inside a PH model in order
to refine its dynamics, and study a more precise class of underlying Thomas models.
%\todo{describe application}

@@

The contributions presented in this paper significantly extend and improve the preliminary results
presented in \cite{FPIMR12-CMSB}.
In addition of improving the efficiency and accuracy of the IG inference, we have added support for
unsigned regulations, that are regulations being positive or negative depending on a particular
context.
This allows to apply our methodology to a wider class of models.
Furthermore, some parts have been completely rewritten in order to explain more precisely the efficiency of the ASP implementation,
and give a new detailed biological application of our work.
